A Decision Procedure for String Constraints with String-Integer Conversion and Flat Regular Constraints

### 2021-05-18 FMCAD2021/3_flatten.tex
section 3
polish

The function  {\parseInt} takes a decimal string as the input and returns the integer represented by the string\footnote{The {\parseInt} function in scripting languages e.g. Javascript is more general in the sense that the base can be a number between 2 and 36. Although our approach works for arbitrary positive bases, we choose to focus on the base 10 in this paper, for readibility.}.
%Since we focus on the decidability,
%we define a binary version of {\parseInt},
%which takes a binary string and returns a decimal integer number.
For example,
${\parseInt}('0123') = {\parseInt}('123')=10^2+10*2+3 = 123$. 
Note here we use the quotation marks to delimit the strings, 
%Clearly,  our decision procedure given in this paper 
%can be adapted to string constraints with other string to number conversion function without substantial change. 
%Single quotation marks are used to distinguish a symbol like $'1'\in \Sigma$
%from a number $1_\mathbb{N}\in \mathbb{N}$ when needed.

Formally, the semantics of the $\parseInt$ function is defined as follows. 
%\begin{definition} 
Let $\Sigma_{\textit{num}}=\{0,1, \ldots, 9\}$. Then ${\parseInt}: \Sigma_{\textit{num}}^+ \mapsto \Nat$ is recursively defined by
    for every $w\in \Sigma_{\textit{num}}^+$,
    \begin{itemize}
%        \item if $|w|=0$, i.e., $w=\epsilon$,  ${\parseInt}(w)=0$;
        \item if $w={'i'}$ for $i \in \Sigma_{\textit{num}}$, then ${\parseInt}('i')=i$;
        \item for $w = w_1 {'i'}$ for $i \in \Sigma_{\textit{num}}$ with $|w_1| \ge 1$, 
        ${\parseInt}(w) = 10*{\parseInt}(w_1)+{\parseInt}({'i'})$.
    \end{itemize} 
%\end{definition}
Note that $\parseInt$ is undefined with $\varepsilon$ as the input.

@@

The flattening of $\varphi \in \strint$ under a flat domain restriction $\flatfun_\varphi$ is a {\paexp} formula, denoted by $\flatten_{\flatfun_\varphi}(\varphi)$, that encodes its flattened semantics.
%
More concretely, $\flatten_{\flatfun_\varphi}(\varphi)$ is a formula over the integer variables $\ivar(\varphi)$,  and Parikh variables $\pvar_{\flatfun_\varphi}(\varphi) = \bigcup_{\svarx \in \svar(\varphi)} \pvar_{\flatfun_\varphi}(\svarx)$, where $\pvar_{\flatfun_\varphi}(\svarx) = \{\#_{\svarx,i} \mid i \in [k_\svarx]\}$, called \emph{flattening variables}, plus some other auxiliary variables, such that 
%
$$\llbracket \phi \rrbracket_\kell =\decode_{\flatfun_\varphi}(\llbracket \flatten_{\flatfun_\varphi}(\varphi) \rrbracket |_{\ivar(\phi) \cup \pvar_{\flatfun_\varphi}(\varphi)})$$
%
The decoding function above decodes an interpretation of integer and flattening variables $I_e: \ivar(\varphi) \cup \pvar_{\flatfun_\varphi}(\varphi) \rightarrow \Nat$ as a set $\decode_{\flatfun_\varphi}(I_e)$ of interpretations of the $\varphi$'s integer and string variables $(I_s, I_i)$ with $I_s: \svar(\varphi) \rightarrow \Sigma^*$ and $I_i: \ivar(\phi) \rightarrow \Nat$ such that 
\begin{itemize}
\item  
for every $ \svarx \in \svar(\varphi)$, $I_s(\svarx) = w_{\svarx,1}^{I_e(\#_{\svarx,1})} \ldots  w_{\svarx,k_\svarx}^{I_e(\#_{\svarx,k_\svarx})}$, 
%
\item for every $ \ivarx \in \ivar(\varphi)$, $I_i(\ivarx) = I_e(\ivarx)$.
\end{itemize}

@@

Before presenting the construction of $\flatten_{\flatfun_\varphi}(\iterm_1 \op \iterm_2)$, for every $\svarx \in \svar(\varphi)$, we define $\flatten_{\flatfun_\varphi}(\parseInt(\svarx)) = \iterm_{\svarx,1}$  such that $(\iterm_{\svarx, i})_{i \in [k_\svarx]}$ and $(\ell_{\svarx, i})_{i \in [k_\svarx]}$ are inductively defined as follows: 
\begin{itemize}
\item for $i = k_\svarx$, 
$$\iterm_{\svarx, i} = \frac{\parseInt(w_{\svarx, k_\svarx}) (10^{|w_{\svarx,k_\svarx}| \#_{\svarx, k_\svarx} } -1 )}{(10^{|w_{\svarx, k_\svarx}|} -1 )}$$ 
and $\ell_{\svarx, i} = |w_{\svarx,k_\svarx}| \#_{\svarx, k_\svarx}$,
%
\item for $i \in [k_\svarx-1]$, 
%
$$\iterm_{\svarx, i} =  \frac{\parseInt(w_{\svarx, i}) (10^{|w_{\svarx, i} | \#_{\svarx, i} } -1 ) 10^{\ell_{\svarx, i+1}}} {(10^{|w_{\svarx, i}|} -1 )} + \iterm_{\svarx, i+1}$$
%
and $\ell_{\svarx, i} = |w_{\svarx, i} | \#_{\svarx, i} + \ell_{\svarx, i+1}$.
\end{itemize}

@@

>>>>

The function  {\parseInt} takes a decimal string as the input and returns the integer represented by the string\footnote{The {\parseInt} function in scripting languages e.g. Javascript is more general in the sense that the base can be a number between 2 and 36. Although our approach works for arbitrary positive bases, we choose to focus on the base 10 in this paper, for readability.}.
%Since we focus on the decidability,
%we define a binary version of {\parseInt},
%which takes a binary string and returns a decimal integer number.
For example,
${\parseInt}('0123') = {\parseInt}('123')=10^2+10*2+3 = 123$. 
Note here we use the quotation marks to delimit the strings, 
%Clearly,  our decision procedure given in this paper 
%can be adapted to string constraints with other string to number conversion function without substantial change. 
%Single quotation marks are used to distinguish a symbol like $'1'\in \Sigma$
%from a number $1_\mathbb{N}\in \mathbb{N}$ when needed.

Formally, the semantics of the $\parseInt$ function is defined as follows. 
%\begin{definition} 
In order to simplify presentation, we assume all string variables ranging over numerical symbols $\Sigma_{\textit{num}}=\{0,1, \ldots, 9\}$. Note that one can easily extend our approach to allow arbitrary finite alphabet. Then ${\parseInt}: \Sigma_{\textit{num}}^+ \mapsto \Nat$ is recursively defined by
    for every $w\in \Sigma_{\textit{num}}^+$,
    \begin{itemize}
%        \item if $|w|=0$, i.e., $w=\epsilon$,  ${\parseInt}(w)=0$;
        \item \yfc{better use `$i$' or $``i"$ instead of $'i'$} if $w={'i'}$ for $i \in \Sigma_{\textit{num}}$, then ${\parseInt}('i')=i$;
        \item for $w = w_1 {'i'}$ for $i \in \Sigma_{\textit{num}}$ with $|w_1| \ge 1$, 
        ${\parseInt}(w) = 10*{\parseInt}(w_1)+{\parseInt}({'i'})$.
    \end{itemize} 
%\end{definition}
Note that $\parseInt$ is undefined with $\varepsilon$ as the input.

@@

The flattening of $\varphi \in \strint$ under a flat domain restriction $\flatfun_\varphi$ is a {\paexp} formula, denoted by $\flatten_{\flatfun_\varphi}(\varphi)$, that encodes its flattened semantics.
%
More concretely, $\flatten_{\flatfun_\varphi}(\varphi)$ is a formula over the integer variables $\ivar(\varphi)$,  and Parikh variables $\pvar_{\flatfun_\varphi}(\varphi) = \bigcup_{\svarx \in \svar(\varphi)} \pvar_{\flatfun_\varphi}(\svarx)$, where $\pvar_{\flatfun_\varphi}(\svarx) = \{\#_{\svarx,i} \mid i \in [k_\svarx]\}$, called \emph{flattening variables}, plus some other auxiliary variables, such that 
%
$$\llbracket \varphi \rrbracket_{\flatfun_\varphi} =\decode_{\flatfun_\varphi}(\llbracket \flatten_{\flatfun_\varphi}(\varphi) \rrbracket |_{\ivar(\phi) \cup \pvar_{\flatfun_\varphi}(\varphi)})$$
%
The decoding function above decodes an interpretation of integer and flattening variables $I_e: \ivar(\varphi) \cup \pvar_{\flatfun_\varphi}(\varphi) \rightarrow \Nat$ as a set $\decode_{\flatfun_\varphi}(I_e)$ of interpretations of the $\varphi$'s integer and string variables $(I_s, I_i)$ with $I_s: \svar(\varphi) \rightarrow \Sigma^*$ and $I_i: \ivar(\phi) \rightarrow \Nat$ such that 
\begin{itemize}
\item  
for every $ \svarx \in \svar(\varphi)$, $I_s(\svarx) = w_{\svarx,1}^{I_e(\#_{\svarx,1})} \ldots  w_{\svarx,k_\svarx}^{I_e(\#_{\svarx,k_\svarx})}$, 
%
\item for every $ \ivarx \in \ivar(\varphi)$, $I_i(\ivarx) = I_e(\ivarx)$.
\end{itemize}

@@

Before presenting the construction of $\flatten_{\flatfun_\varphi}(\iterm_1 \op \iterm_2)$, for every $\svarx \in \svar(\varphi)$, we define $\flatten_{\flatfun_\varphi}(\parseInt(\svarx)) = \iterm_{\svarx,1}$  such that $(\iterm_{\svarx, i})_{i \in [k_\svarx]}$ and $(\ell_{\svarx, i})_{i \in [k_\svarx]}$ are inductively defined as follows: 
\begin{itemize}
\item for $i = k_\svarx$, 
$$\iterm_{\svarx, i} = \frac{\parseInt(w_{\svarx, k_\svarx}) (10^{|w_{\svarx,k_\svarx}| \#_{\svarx, k_\svarx} } -1 )}{(10^{|w_{\svarx, k_\svarx}|} -1 )}$$ 
and $\ell_{\svarx, i} = |w_{\svarx,k_\svarx}| \#_{\svarx, k_\svarx}$,
%
\item for $i \in [k_\svarx-1]$, 
%
$$\iterm_{\svarx, i} =  \frac{\parseInt(w_{\svarx, i}) (10^{|w_{\svarx, i} | \#_{\svarx, i} } -1 ) 10^{\ell_{\svarx, i+1}}} {(10^{|w_{\svarx, i}|} -1 )} + \iterm_{\svarx, i+1}$$
%
and $\ell_{\svarx, i} = |w_{\svarx, i} | \#_{\svarx, i} + \ell_{\svarx, i+1}$.
\end{itemize}
Notice that here $|w_{\svarx, \textunderscore } |$ and $\ell_{\svarx, \textunderscore}$ are constants while $\#_{\svarx, \textunderscore}$ are (flattening) variables.

@@

%The correctness of the flattening approach is stated below.
%\begin{proposition}
%$\llbracket \varphi \rrbracket_{\flatfun_\varphi}$ is nonempty iff $\flatten_{\flatfun_\varphi}(\varphi)$ is satisfiable.
%\end{proposition}
